---
title: "et_eller_andet"
format: html
editor: visual
---

## display data

```{r}
#| message: false
library("tidyverse")
library("viridis")
```

```{r}
#| warning: false
#| message: false
data <- read_tsv("./../data/03_dat_aug.tsv")

data |>
  group_by(case_control,age_visit,timepoint)|>
  summarize(total_temp_mu=log(mean(total_rearrangements))) |>
  ggplot(data=_,
         mapping=aes(x=age_visit,y=total_temp_mu))+
  geom_point(aes(color=timepoint))+
  labs(title="mean of total rearrangements vs. age at visit")+
  scale_color_viridis()+
  geom_smooth(method="lm",color="black")+
  facet_grid(vars(case_control))

data |>
  group_by(case_control,age_visit,timepoint)|>
  summarize(prod_temp_mu=log(mean(productive_rearrangements))) |>
  ggplot(data=_,
         mapping=aes(x=age_visit,y=prod_temp_mu))+
  geom_point(aes(color=timepoint))+
  labs(title="mean of productive rearrangements vs. age at visit")+
  scale_color_viridis()+
  geom_smooth(method="lm",color="black")+
  facet_grid(vars(case_control))


data |>
  group_by(case_control,age_visit,timepoint)|>
  summarize(frac_prod=mean(productive_rearrangements)/mean(total_rearrangements)) |>
  ggplot(data=_,
         mapping=aes(x=age_visit,y=frac_prod))+
  geom_point(aes(color=timepoint))+
  labs(title="fraction of productive rearrangements vs. age at visit")+
  scale_color_viridis()+
  geom_smooth(method="lm",color="black")+
  facet_grid(vars(case_control))
```

We see a slight rise in total and productive rearrangements for case patients and a decrease for control patients, when comparing the values against age at visit. So the number of rearrangements seems to be stable in case patients during aging, however there is decreasing rearrangement for the control patients.

We see a steep rise in the fraction of productive rearrangement with age, which must mean that the differences in values between productive rearrangement and total rearrangement become smaller with age.

```{r}
#simpson clonality plot

data |> 
  ggplot(data=_,
         mapping=aes(x=age_visit,y=log(productive_simpson_clonality)))+
  geom_point(aes(color=timepoint))+
  labs(title="mean of total rearrangements pr. timepoint")+
  scale_color_viridis()+
  geom_smooth(method="lm",color="black")+
  facet_grid(~case_control)

data|>
  filter(case_control=="Case")|>
  lm(productive_simpson_clonality~age_visit, data=_)

data|>
  filter(case_control=="Control")|>
  lm(productive_simpson_clonality~age_visit, data=_)
```

Both case and control show a direct proportionality between productive simpson clonality and age, which shows the decrease of polyclonality in the individuals.

```{r}

TCR_tidy <- data |> 
  select(timepoint, age_visit,
         starts_with('HLA'), GAD65, case_control,gender,age) |>
  arrange(desc(desc(timepoint))) |> 
  pivot_longer(starts_with('HLA_'), 
               names_to = "allele",
               values_to = "subtype") |>
  mutate(allele = str_remove(allele, "(\\,1)|(\\,2)")) |>
  group_by(allele, subtype) |> 
  mutate(allele_count = n(), .groups = NULL) |>
  filter(subtype != "NA")


ggplot(data = TCR_tidy) +
  geom_tile(mapping = aes(x = allele,
                          y = subtype,
                          fill = allele_count)
            ) + 
  theme(axis.text.y = element_text(size = 4),
        axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient(low = "#3333FF",
                    high = "#FF3300")+
  facet_grid(~gender)
```

```{r}
data |>
  ggplot() +
  geom_histogram(aes(x = gender),stat="count",
               size = 0.5, fill = "darkred",
               bins = 15, color = "black") +
  labs(title = "Age of diagnosis, type 1 diabetes",
       subtitle = "Histogram",
       x = "Age of diagnosis")+
  facet_grid(~case_control)

data |>
  ggplot(data=_,
         mapping=aes(x=age_visit,y=IAA))+
  geom_point(aes(color=case_control))+
  facet_grid(~case_control)+
  scale_color_viridis(discrete=TRUE)

```

PCA Alleles

```{r}
HLA_tidy <- data |> 
  select(starts_with('HLA'), case_control,id,timepoint) |> 
  pivot_longer(starts_with('HLA_'), 
               names_to = "allele",
               values_to = "subtype") |>
  mutate(allele = str_remove(allele, "(\\,1)|(\\,2)")) |>
  group_by(allele,case_control,id,timepoint) |> 
  mutate(allele_count = n(), .groups = NULL) |>
  filter(subtype != "NA") |>
  pivot_wider(names_from = allele, values_from=allele_count)
  
```

```{r}
library("broom")
library("cowplot")

data <- read_tsv("./../data/03_dat_aug.tsv",na="N/A")
#biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")

pca_data <- data|>
  select(case_control) |>
  select(where(is.numeric))|>
  scale()|>
  prcomp()

  
TCR_tidy <- data |> 
  select(timepoint, age_visit,
         starts_with('HLA'), GAD65, case_control) |>
  arrange(desc(desc(timepoint))) |> 
  pivot_longer(starts_with('HLA_'), 
               names_to = "allele",
               values_to = "subtype") |>
  mutate(allele = str_remove(allele, "(\\,1)|(\\,2)")) |>
  group_by(allele, subtype) |> 
  mutate(allele_count = n(), .groups = NULL) |>
  filter(subtype != "NA")

pca_data |>
  augment(pca_data) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#D55E00", benign = "#0072B2")
  ) +
  theme_half_open(12) + background_grid()

pca_fit <- data %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>% # scale data
  prcomp() # do PCA



data <- read_tsv("./../data/03_dat_aug.tsv")|>
  mutate(across(everything(), 
                ~replace_na(., 0))) 

data_pca <- read_tsv("./../data/03_dat_aug.tsv")|>
  mutate(across(everything(), 
                ~replace_na(., 0))) |>
  select(case_control, starts_with("HLA")) |>
  select(where(is.numeric)) |> # retain only numeric columns
  scale() |> # normalizing the data 
  prcomp()


# plotting PC coordinates for individuals
data_pca |>
  augment(data) |># add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, 
             color = case_control)) + 
  geom_point(size = 1.5) +
  scale_color_viridis(discrete=TRUE)+
  theme_half_open(12) + 
  background_grid() +
    labs(title = "Plot of individuals",
         color = "Case vs Control")

# plotting scree plot for varialbe variation
data_pca |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#0072B2", 
           alpha = 0.8) + # making the bars slightly translusent 
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))) +
  geom_line(aes(PC),
            color = "red") +
  theme_minimal_hgrid(12) +
  background_grid() +

  labs(title = "Scree Plot",
       y = "percentage contribution")

# plotting the rotation matrix
arrow_style <- arrow(length = unit(0.2, "cm"), 
                     ends = "first")

data_pca |>
  tidy(matrix = "rotation") |>
  pivot_wider(names_from = "PC", 
              names_prefix = "PC", 
              values_from = "value") |>
  ggplot(aes(x = PC2, y = PC1)) +
  geom_point(alpha = 0) +  # making the points fully transparent
  geom_segment(aes(xend = 0, yend = 0), 
               arrow = arrow_style, 
               color = "black") +
  coord_flip() +
  geom_text(aes(label = column), hjust = 1.2, nudge_x = -0.02, color = "#0072B2") +
  theme_half_open(12) + 
  background_grid() +
  labs(title = "Rotation Matrix")
```
