---
<<<<<<< HEAD
title: "Render all files"
=======
title: "99_render_all"
>>>>>>> b4f2d4812c1baee66bc9defd55ee4a4efce44de5
format: html
editor: visual
---

<<<<<<< HEAD
=======
Rendering files

>>>>>>> b4f2d4812c1baee66bc9defd55ee4a4efce44de5
```{r}
library("quarto")
quarto_render("00_all.qmd")
```

<<<<<<< HEAD
```{bash}
ls ./../results/*.html | xargs -I{} rm {}
ls ./*.html | xargs -I{} mv {} ./../results/
=======
Moving the files:

```{r}
file.rename(from = "./00_all.qmd",
            to = "./../00_all.qmd")
file.rename(from = "./01_load.qmd",
            to = "./../01_load.qmd")
file.rename(from = "./02_clean.qmd",
            to = "./../02_clean.qmd")
file.rename(from = "./03_augment.qmd",
            to = "./../03_augment.qmd")
file.rename(from = "./04_describe.qmd",
            to = "./../04_describe.qmd")
file.rename(from = "./05_antibody_expression_analysis.qmd",
            to = "./../05_antibody_expression_analysis.qmd")
file.rename(from = "./06_pca_antibody.qmd",
            to = "./../06_pca_antibody.qmd")
file.rename(from = "./07_HLA_allele_analysis.qmd",
            to = "./../07_HLA_allele_analysis.qmd")
file.rename(from = "./08_pca_HLA.qmd",
            to = "./../08_pca_HLA.qmd")
>>>>>>> b4f2d4812c1baee66bc9defd55ee4a4efce44de5
```
